---
title: "Math Student Alcohol Consumption"
date: "March 15, 2017"
output: html_document
runtime: shiny
---
## Contributors
* Bryan Daetz
* Daniel Zeng
* Moritz Droste
* Jiali Huang

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About the Data Set
In our analysis of the Student Alcohol Consumption data set, we chose to look at only the math students. Our graphs and analysis focus on levels of alcohol consumption on a scale of 1 (low) to 5 (high) and how it relates to varying demographics categories.

## Importing Packages

```{r, message = FALSE}
#importing necessary packages
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyr)
library(stringr)
```

## Cleaning Data Set

```{r}
#importing data set (Student Alchol Consumption) for math students only
math <- read.csv("C://Users/bdaet/Desktop/R_Tutorials/student-mat.csv")

#creating a variable combining weekday and weekend alcohol consumption
math2 <- mutate(math, Total_Consumption = math$Dalc + math$Walc)

#some of the variables were interpreted as numerics when they are actually factors so we have to change them to factors
math$Dalc <- as.factor(math$Dalc)
math$Walc <- as.factor(math$Walc)
math$age <- as.factor(math$age)

#the failures variable is set up weirdly, it's numeric from 1-3 and then 4 means 4 or more
#to compensate for this I'm going to change it to a factor and change 4 to 4+ to avoid confusion
math$failures <- as.factor(math$failures)
math$failures <- str_replace(math$failures, pattern = "4", replacement = "4+")
#in this data set it turns out there are no students who failed 4+ classes so this wasn't actually necessary

math <- gather(math, Time_of_Week, Alc_Consumption, Dalc, Walc)
math$Time_of_Week <- str_replace(math$Time_of_Week, pattern = "Dalc", replacement = "Weekday")
math$Time_of_Week <- str_replace(math$Time_of_Week, pattern = "Walc", replacement = "Weekend")

```

## Alcohol Consumption by Gender

```{r, fig.height = 6, fig.width = 10}

#looking at weekday alcohol consumption by gender
mvf <- math %>%
          group_by(sex, Time_of_Week, Alc_Consumption) %>%
          summarise(Count = n())

male_v_female <- ggplot(mvf, aes(x = Alc_Consumption, y = Count, fill = sex)) +
                      geom_bar(stat = "identity", position = "dodge", alpha = 0.85) +
                      theme_bw() +
                      facet_wrap(~ Time_of_Week) +
                      scale_fill_manual(values = c("hotpink","royalblue")) +
                      ggtitle("Alcohol Consumption by Gender") +
                      xlab("Consumption Level (1 = Very Low, 5 = Very High)")
ggplotly(male_v_female)
```

One of the most apparent trends can be observed in the graph "Alcohol Consumption By Gender". At low alcohol consumption levels, there is a higher percentage of female drinkers than males drinkers. Specifically, 40.1% of females drink only low levels of alcohol during the weekdays compared to males, where 29.1% drink at low levels during weekdays. On the weekends, a similar trend can be observed, with 22.6% of females drinking at low levels compared to 15.2% of males. When looking at high levels of alcohol consumption, there is a higher percentage of male drinkers (2.1%) compared to female drinkers (0.24%) during weekdays, as well as weekends, with 6.4% and 0.96% respectively.

## Alcohol Consumption by Age

```{r, fig.height = 6, fig.width = 10}
#looking at alcohol consumption by age
#after looking at the result, it appears there is little to no data on ages 20-22 in this data set
#to make a more readable graph we're only going to focus on ages 15-19
ages <- math %>%
          filter(age %in% c(15:19)) %>%
          group_by(age, Alc_Consumption, Time_of_Week) %>%
          summarise(Count = n())

a <- ggplot(ages, aes(x = Alc_Consumption, y = Count, fill = Time_of_Week)) +
          geom_bar(stat = "identity", alpha = 0.65, position = "dodge") +
          theme_bw() +
          facet_wrap(~ age) +
          scale_fill_manual(values = c("skyblue", "limegreen")) +
          ggtitle("Alcohol Consumption by Age") +
          xlab("Consumption Level (1 = Very Low, 5 = Very High)") 
ggplotly(a)
```

When observing variations in alcohol consumption levels based on age, according to the graph "Alcohol Consumption By Age", there is a larger number of low-consumption during weekdays than weekends, but a larger number of high-consumption during weekends than weekdays, regardless of age. Age does not seem to affect trends in consumption levels throughout the week.

## Extra-Curricular Activities and Alcohol Consumption

```{r, fig.height = 6, fig.width = 10}
#let's look at if participation in extra-curricular activities affects alcohol consumption during the week
extra_act <- math %>%
              group_by(activities, Time_of_Week, Alc_Consumption) %>%
              summarise(count = n())

ext <- ggplot(extra_act, aes(x = Alc_Consumption, y = count, fill = activities)) +
          geom_bar(stat = "identity", position = "dodge", alpha = 0.85) +
          theme_bw() +
          facet_wrap(~ Time_of_Week) +
          scale_fill_manual(values = c("green", "orange")) +
          ggtitle("Effect Participation in Extra-Curricular Activities Has on Alcohol Consumption") +
          xlab("Consumption Level (1 = Very Low, 5 = Very High)")
ggplotly(ext)
```

When looking at the graph "Effect Participation in Extra-Curricular Activities has on Alcohol Consumption", it can be observed that participating in extra-curriculars does not seem to have any particular effect on levels of alcohol consumption.

## Absences and Alcohol Consumption

```{r, fig.height = 6, fig.width = 10}

#let's see if alcohol consumption leads to more absences from class
absences <- ggplot(math, aes(x = Alc_Consumption, y = absences)) +
               geom_boxplot(alpha = 0.65, color = "red4") +
               theme_bw() +
               facet_wrap(~ Time_of_Week) +
               ggtitle("Effect Alcohol Consumption has On Absences from Class") +
               xlab("Consumption Level (1 = Very Low, 5 = Very High)") +
               ylab("Number of Absences")
ggplotly(absences)
```

Next is the graph "Effect Alcohol Consumption has on Absences from Class": The graph tells us that there is a slight positive correlation between level of Alcohol Consumption and Number of Absences. This trend is consistent for both weekdays and weekends, although this relationship is slightly weaker for drinking during the weekends.

## Failed Classes and Alcohol Consumption

```{r, eval = FALSE}

#let's see if weekday alcohol consumption leads to more failed classes
fail <- math %>%
        filter(Time_of_Week == "Weekday") %>%
        group_by(Alc_Consumption, failures) %>%
        summarise(Count = n())

fail <- spread(fail, failures, Count, fill = 0)
names(fail) <- c("Alcohol Consumption Level", "0 Failed Classes", 
                 "1 Failed Class", "2 Failed Classes", "3 Failed Classes")
fail
```

|Alcohol Consumption Level | 0 Failed Classes | 1 Failed Class | 2 Failed Classes | 3 Failed Classes |
|--------------------------|------------------|----------------|------------------|------------------|
|                1         |       232        |       29       |           7      |            8     |
|                2         |        50        |       12       |           8      |            5     |
|                3         |        17        |        6       |           1      |            2     |
|                4         |         6        |        2       |           1      |            0     |
|                5         |         7        |        1       |           0      |            1     |


## Grades and Alcohol Consumption

```{r, fig.height = 6, fig.width = 10}
                
#let's see if alcohol consumption has any effect on final course grades
grades <- ggplot(math, aes(x = Alc_Consumption, y = G3)) +
              geom_boxplot(alpha = 0.65, color = "navy") +
              theme_bw() +
              facet_wrap(~ Time_of_Week) +
              ggtitle("Effect Alcohol Consumption has On Grades") +
              xlab("Consumption Level (1 = Very Low, 5 = Very High)") +
              ylab("Final Grade (on scale from 0-20)")
ggplotly(grades)
```

Our last graph was "Effect Alcohol Consumption has on Grades". The grading scale used at these secondary schools was given on a discrete scale from 0 to 20, which was foreign to us. However, it can still be seen on the graph that there is little to no correlation between alcohol consumption levels and final grades for Math.

```{r, fig.height = 6, fig.width = 10}
scattermath <- ggplot(math2, aes(Total_Consumption,G3)) + geom_jitter(aes(colour = goout)) +
  geom_smooth(color = "red") +
  ggtitle("Total Alcohol Consumption vs Final Math Grades") +
  labs(x = "Total Alcohol Consumption (Rating scaled out of 10)", 
       y = "Final Math Grade (Scaled out of 20)")

ggplotly(scattermath)
```

Combining the weekday and weekend alcohol consumption levels into one score (out of 10) still did not reveal any significant correlation between alcohol consumption and grades.